`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi

### 2023-04-06 manuscript/manuscript.tex
Render

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Habitat International}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi},
  pdfauthor={Elizabeth Tilley; Hope Chilunga; Jonathan Kwangulero; Lars Schöbitz; Heiko Heilgendorff; Marc Kalina},
  pdfkeywords={waste management, healthcare waste, trash burning, air
quality, Malawi, urbanisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{`It Is Unbearable To Breath Here': Air Quality, Open
Incineration, And Misinformation In Blantyre, Malawi}
\author[1]{Elizabeth Tilley%
\corref{cor1}%
}
 \ead{tilleye@ethz.ch} 
\author[2]{Hope Chilunga%
%
}
 \ead{hchilunga@poly.ac.mw} 
\author[]{Jonathan Kwangulero%
%
}
 \ead{kwangerlo@gmail.comm} 
\author[1]{Lars Schöbitz%
%
}
 \ead{lschoebitz@ethz.ch} 
\author[3]{Heiko Heilgendorff%
%
}
 \ead{me@me.com} 
\author[1]{Marc Kalina%
%
}
 \ead{mkalina@ethz.ch}

@@

\begin{abstract}
Blantyre, Malawi's Queen Elizabeth Central Hospital (QECH), or Queen's,
as it's known locally, is the country's largest public hospital.
However, Queen's is not served by regular municipal waste collection.
What municipal collection that is done is ad-hoc, sporadic, and based on
the hospital's available financial resources. Rather, most hospital
waste (infectious and non-infectious) is gathered by grounds staff and
openly burned, in several constantly smouldering piles, sending up
clouds of smoke. Speaking directly to an identified knowledge gap on air
quality impacts linked to trash burning and the paucity of African urban
dwellers' voices on air quality issues, this study employed a
mixed-methods approach to both quantitatively measure the air quality
around QECH, and to qualitatively investigate the perceived impacts
amongst staff and caregivers. Low-cost sensors measuring particulate
matter (PM) with particle sizes less than 10 µm (PM\textsubscript{10})
and less than 2.5 µm (PM2.5), expressed as the mass of PM per volume of
air (μg PMx/m3 air) were recorded every 5 minutes at 8 locations across
the QECH for 2 months. Qualitative data collection consisted of 57
interviews with patients, caregivers and hospital staff (including
janitorial and maintenance staff, nurses, doctors, and administrators).
Our results show that safe air quality thresholds are consistently
exceeded across space and time and that the most problematic air quality
surrounds the shelter for caregivers and those receiving treatment for
HIV/AIDs. Moreover, staff and visitors are severely impacted by the poor
air quality within the space, but feel powerless to make changes or
address complaints. Waste management interventions are desperately
needed lest the patients who arrive at Queen's leave with more health
issues than the ones with which they arrived.
\end{abstract}

\begin{highlights}
\item Low-cost monitors were used to assess air quality at Malawi's
largest hospital\item Open waste burning is physically extensive and
temporally frequent across the hospital campus\item Air quality measured
as PM2.5 and PM10 consistently exceeded even the most hazardous
levels\item The most exposed locations were where guardians of patients
stay and the HIV clinic
\end{highlights}

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, frame hidden, breakable]}{\end{tcolorbox}}\fi

@@

Poor air quality in in African cities can be attributed to multiple
factors, including vehicular emissions, dust, or the use of low-quality
fuels for cooking, lighting, and heating, however, the open burning of
waste has been identified as one of the key contributors to air
pollution within urban areas
\citep{amegah2018proliferation, bulto2020impact}. Although open trashing
burning is a problem in LICs broadly, for instance,
\citet{wiedinmyer2014global} estimate that open trash burning
contributes to 29\% of the total global PM2.5 anthropogenic emissions,
it is of particular concern in Sub-Saharan African (SSA) cities, which
are home to 19 of the world's 50 biggest dumpsites (United Nations
Development Programme, 2021). Particulate matter, or PM, is not specific
to trash burning, but rather the product of any type of incomplete
combustion and is one of the most commonly measured indicators for air
quality. PM is measured in terms of its particle size and generally
classified into PM with particle sizes less than 10 µm (PM10) and those
with sizes less than 2.5 µm (PM2.5), while one way of measuring the
severity of air pollution is by expressing the mass of PM per volume of
air (μg PMx/m3 air). PM has been linked to multiple negative health
outcomes. Amongst particulate air pollution, PM10 and PM2.5 are of
particular concern as they can penetrate deep into the lungs, and their
exposure has been associated with asthma, heart disease, heart failure,
stroke, and cancer \citep{mohapatra2014effect, xing2016impact}. As a
result, the health impacts of ambient air particle pollution may be
significant, especially for the most vulnerable. For instance, the
\citet{ihme2019ambient} estimates PM related air pollution resulted in
4.14 million deaths worldwide (in 2019), of which, 394,000 occurred in
Africa. Yet, despite its importance, the air quality and health picture
in Africa is worryingly incomplete. For instance, the WHO only collects
data from 10 African countries (covering 39 cities)
\citep{worldhealthorganization2016ambient}. Likewise, robust,
epidemiological data from and for the African context is also lacking,
though emerging data suggests strong correlations between PM and poor
health outcomes \citep{coker2018narrative, heft-neal2018robust}. Moving
forward, it is not just the air at Queen's which lacks clarity.

@@

Findings suggest that particulate matter concentrations are routinely
above the WHO guidelines and are especially worrisome at several
locations. Over the course of the two-month study, the hazardous limits
for both PM10 and PM2.5 were exceeded at all locations except for the
Lions Sight hospital. The limits were exceeded fewer than 50 times at
five of the locations, but the monitors at the Lighthouse Clinic and the
Guardian Shelter recorded more than 50 instances above the hazardous
limits. The extremely hazardous air quality at the Guardian shelter is
mostly from cooking over open fires, while the PM at the Lighthouse
Clinic is a function of being located directly adjacent to the
smouldering waste pile.

@@

Yet, according to respondents and our own observations, waste management
procedures within the hospital are fraught with challenges which impact
the implementation of best practices. These include the lack of a
coherent hospital-wide waste management policy, insufficient waste
management training for staff, and a persistent lack of material and
financial resources. For instance, although hospital administration was
able to articulate a set of standard operating principles (SoPs) for
waste management, below the administration, doctors and nurses were
generally limited in their understanding about its contents.
Furthermore, cleaners, who are responsible for day-to-day cleansing and
waste disposal, yet rate at the bottom of the institutional hierarchy at
Queens, were unable to access or analyse hospital policy, and generally
tended to rely on ``experience'' and ``human judgement'' when handling
potentially hazardous waste. Moreover, although nearly all doctors and
nurses had prior education on waste management best practices, cleaners
began their waste handling duties without any prior waste management
experience, and although there is an orientation programme for new
hires, there is no regular refresher training or capacity building
regarding waste for existing staff, at any level.

@@

The PM monitoring device centres around a Raspberry Pi 3 Model B
single-board computer complete with an operating system and storage
space. The added advantage of storage space is that the data are not
lost if the internet connection breaks down. Because the Pi can only
read digital signals, we needed to include an analog-to-digital (ADC)
converter. The amount of dust is measured by a Nova PM SDS011 High
Precision Laser sensor. This sensor measures particles at 2.5 and 10
micrometres in diameter in mg/m3. A list of the specific hardware and
software components can be found in Appendix B.

@@

@@

The Malawi Bureau of Standards is the national agency responsible for
setting and publishing all standards in the country. At the time of
writing, the official website (www.mbsmw.org) was not available.
However, published work that references Malawian standards
\citep{kutlarjoss2017time, mapoma2014air} indicate that the maximum 24-h
PM10 value is 25 μg/m3 and the maximum annual PM2.5 value is 8 μg/m3.
There is no daily maximum value for PM2.5. It should be noted that both
of these values are at, or below, the level-4 interim WHO targets.

@@

\hypertarget{tbl-peaks}{}
\begin{longtable}[]{@{}lrrr@{}}
\caption{\label{tbl-peaks}Peaks for both PM\textsubscript{2.5} and
PM\textsubscript{10} are the number of data points above the WHO interim
target 1 (annual) which is the least stringent; the number of
observations recorded (count) is provided for reference.}\tabularnewline
\toprule()
location & n & pm10 & pm2.5 \\
\midrule()
\endfirsthead
\toprule()
location & n & pm10 & pm2.5 \\
\midrule()
\endhead
Administration & 10900 & 171 & 94 \\
Guardian Shelter & 11032 & 4529 & 5314 \\
Lighthouse Clinic & 11472 & 4365 & 3891 \\
Lions Sight & 11749 & 91 & 14 \\
Malaria & 12031 & 617 & 578 \\
Mercy James & 12193 & 736 & 749 \\
Ward 6B & 11861 & 1113 & 1236 \\
uMoyo & 10637 & 703 & 399 \\
\bottomrule()
\end{longtable}

Over the course of the two-month study, the hazardous limits for both
parameters were exceeded at all locations except for the Lions Sight
hospital. The limits were exceeded fewer than 50 times at five of the
locations, and only the monitors at the Lighthouse Clinic and the
Guardian Shelter recorded more than 50 instances above the hazardous
limits. At the Lighthouse Clinic, the PM10 limit was exceeded almost
twice as often as the PM2.5 limit, while at the Guardian Shelter, the
reverse is true. Because the Lighthouse Clinic is within 50 m of the
incinerator, the number and predominance of PM10 peaks is characteristic
of incomplete combustion and dust that is typical for the area and the
open burning that occurs there. The residents at the Guardian Shelter
however are mostly cooking within contained clay stoves that have been
designed to improve combustion of the coal and wood that is burned
within them. The fewer number of PM10 peaks is a testament to this
intervention, though the frequency of PM2.5 peaks is still beyond
acceptable.

@@

The 24-hour averaged values (logged every 5 minutes) are presented for
each location, for both PM10 and PM2.5 in fig-24-hour-average. As well
as dampening local variation and peaks, the WHO air quality guidelines
are also based on 24-hour averages which are the standard against which
the health risks can be judged.

Overall, PM2.5 values remained below 100 μg/m3 at 6 of 8 locations
(Administration, Lions Sight, Malaria Project, uMoyo, Ward 6B, and Mercy
James); PM10 values were consistently below 100μg/m3 at the same
locations, but with several average values extending slightly above, and
then infrequently.

@@

\caption{\label{fig-24-hour-average}Average 24-hour PM10 and PM2.5 at 8
monitoring stations over 2 months}

@@

Given that the type of fuel and the type of burning (contained vs.~open)
produce very different particulate ``fingerprints'', the ratio of PM10
to PM2.5 values were examined to determine if a clear difference between
locations, and therefore source, could be identified. The results are
presented in Figure~\ref{fig-pm-ratios}.

@@

\caption{\label{fig-pm-ratios}PM10:PM2.5 values for each location. Each
line shows the minimum and maximum ratio for the day based on ratios
calculated hourly. Does not include 14th to 16th October at location
uMoyo due to extreme outliers.}

@@

The results presented in Figure 5 illustrate both the presence of PM10
relative to PM2.5 as well as the range of values over which that ratio
is observed.

Unlike the peak values presented in Table 2, the values presented in
Figure 5 indicate the relative presence of larger (PM10) particles to
smaller ones (PM2.5) regardless of their concentration: the ratio of two
low concentrations can be the same as the ratio of two high
concentrations and is therefore more indicative of source than of
proximity.

@@

The air quality, as a result of open waste burning at Queen's is poor
and not suitable for a city, let alone a hospital. Over the course of
this two-month study, the hazardous limits for both PM parameters were
exceeded at all locations except for the Lions Sight hospital. The WHO
limits were exceeded fewer than 50 times at five of the locations, and
only the monitors at the Lighthouse Clinic and the Guardian Shelter
recorded more than 50 instances above the most hazardous limit. The
composition of emissions recorded ranged significantly, analysed as the
ratio PM10:PM2.5 likely reflects a range of burning styles and/or fuel
(waste) type, and no discernable trend was observed.. Among interview
respondents, there was a general consensus that air quality caused by
the burning of waste within the hospital was a problem, although there
were significant differences between respondents over their
understandings of potential impacts, the effectiveness of various coping
mechanisms, and their problematisations linked to the burning of
specific waste materials. For most, going home or leaving the Queen's
campus was the only pathway to relief, though for patients who can not
leave, endurance was the only option. This work was originally planned
as a before-and-after study: we envisioned reporting these baseline
results compared to follow-up measurements once the long-promised
incinerator was commissioned. However, nearly 4 years later, the
incinerator is rarely functional (partly for technical reasons, partly
for financial ones) and the smoke hovering over Queen's is still thick.
Waste management has long been known as an environmental health issue,
but in the case of open trash burning at a medical facility, it becomes
more: for some of the most vulnerable, especially those with TB, asthma
or other chronic disease, it could be a matter of life or death.

@@

\caption{\label{fig-five-minute-intervals}PM2.5 and PM10 values
collected every 5 minutes over 10 months at 8 locations.}

@@

\begin{itemize}
\tightlist
\item
  analog-to-digital converter (ADC)
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  any box that can fit all the components comfortbaly can be used. The
  box will need to have a whole drilled in it for the pipe.
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The Raspberry Pi operating system runs off of Rasbian software on a
  NOOBS scandisk.
\item
  Air quality monitoring code is written in Python using the
  \href{https://pypi.org/project/python-aqi/}{aqi library} (A library of
  algorithms to convert between AQI value and pollutant concentration).
  The code is stored on the
  \href{https://github.com/opendatadurban/hospital_stations}{Open Data
  Durban GitHub account} and is free for all to access. To be able to
  collect data, the device needs to be connected to a reliable wifi
  network, the SSID and password need to be coded onto the device during
  step 2 in the building method.
\item
  Once the whole device has been put together and placed on site, to
  access that data we use a combination of propgrams:
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set up the Raspberry Pi with the Rasbian software using the NOOBS
  scandisk. (insert the scandisk into the Rasberry Pi) A comprehensive
  start up guide can be found here.
\item
  Next we load the code onto the Raspberry Pi. The code can be found on
  the Open Data Durban Hospital Stations Repository Github. To use it
  you can follow the steps in the file called Initiate\_WS.txt. Once all
  the steps have been completed, the Raspberry Pi is now a device that
  can be connected to a particulate matter monitoring sensor.\\
\item
  Connect the hardware components
\end{enumerate}

>>>>

\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\journal{Habitat International}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={`It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi},
  pdfauthor={Elizabeth Tilley; Hope Chilunga; Jonathan Kwangulero; Lars Schöbitz; Saloni Vijay; Heiko Heilgendorff; Marc Kalina},
  pdfkeywords={waste management, healthcare waste, trash burning, air
quality, Malawi, urbanisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{`It Is Unbearable To Breath Here': Air Quality, Open
Incineration, And Misinformation In Blantyre, Malawi}
\author[1]{Elizabeth Tilley%
\corref{cor1}%
}
 \ead{tilleye@ethz.ch} 
\author[2]{Hope Chilunga%
%
}
 \ead{hchilunga@poly.ac.mw} 
\author[]{Jonathan Kwangulero%
%
}
 \ead{kwangerlo@gmail.comm} 
\author[1]{Lars Schöbitz%
%
}
 \ead{lschoebitz@ethz.ch} 
\author[1]{Saloni Vijay%
%
}
 \ead{svijay@ethz.ch} 
\author[3]{Heiko Heilgendorff%
%
}
 \ead{me@me.com} 
\author[1]{Marc Kalina%
%
}
 \ead{mkalina@ethz.ch}

@@

\begin{abstract}
Blantyre, Malawi's Queen Elizabeth Central Hospital (QECH), or Queen's,
as it's known locally, is the country's largest public hospital.
However, Queen's is not served by regular municipal waste collection.
What municipal collection that is done is ad-hoc, sporadic, and based on
the hospital's available financial resources. Rather, most hospital
waste (infectious and non-infectious) is gathered by grounds staff and
openly burned, in several constantly smouldering piles, sending up
clouds of smoke. Speaking directly to an identified knowledge gap on air
quality impacts linked to trash burning and the paucity of African urban
dwellers' voices on air quality issues, this study employed a
mixed-methods approach to both quantitatively measure the air quality
around QECH, and to qualitatively investigate the perceived impacts
amongst staff and caregivers. Low-cost sensors measuring particulate
matter (PM) with particle sizes less than 10 µm (PM\textsubscript{10})
and less than 2.5 µm (PM\textsubscript{2.5}), expressed as the mass of
PM per volume of air (µg PMx/m\textsuperscript{3} air) were recorded
every 5 minutes at 8 locations across the QECH for 2 months. Qualitative
data collection consisted of 57 interviews with patients, caregivers and
hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators). Our results show that safe air quality
thresholds are consistently exceeded across space and time and that the
most problematic air quality surrounds the shelter for caregivers and
those receiving treatment for HIV/AIDs. Moreover, staff and visitors are
severely impacted by the poor air quality within the space, but feel
powerless to make changes or address complaints. Waste management
interventions are desperately needed lest the patients who arrive at
Queen's leave with more health issues than the ones with which they
arrived.
\end{abstract}

\begin{highlights}
\item Low-cost monitors were used to assess air quality at Malawi's
largest hospital\item Open waste burning is physically extensive and
temporally frequent across the hospital campus\item Air quality measured
as PM\textsubscript{2.5} and PM\textsubscript{10} consistently exceeded
even the most hazardous levels\item The most exposed locations were
where guardians of patients stay and the HIV clinic
\end{highlights}

\begin{keyword}
    waste management \sep healthcare waste \sep trash burning \sep air
quality \sep Malawi \sep 
    urbanisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, boxrule=0pt, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden]}{\end{tcolorbox}}\fi

@@

Poor air quality in in African cities can be attributed to multiple
factors, including vehicular emissions, dust, or the use of low-quality
fuels for cooking, lighting, and heating, however, the open burning of
waste has been identified as one of the key contributors to air
pollution within urban areas
\citep{amegah2018proliferation, bulto2020impact}. Although open trashing
burning is a problem in LICs broadly, for instance,
\citet{wiedinmyer2014global} estimate that open trash burning
contributes to 29\% of the total global PM\textsubscript{2.5}
anthropogenic emissions, it is of particular concern in Sub-Saharan
African (SSA) cities, which are home to 19 of the world's 50 biggest
dumpsites (United Nations Development Programme, 2021). Particulate
matter, or PM, is not specific to trash burning, but rather the product
of any type of incomplete combustion and is one of the most commonly
measured indicators for air quality. PM is measured in terms of its
particle size and generally classified into PM with particle sizes less
than 10 µm (PM\textsubscript{10}) and those with sizes less than 2.5 µm
(PM\textsubscript{2.5}), while one way of measuring the severity of air
pollution is by expressing the mass of PM per volume of air (µg
PMx/m\textsuperscript{3} air). PM has been linked to multiple negative
health outcomes. Amongst particulate air pollution, PM\textsubscript{10}
and PM\textsubscript{2.5} are of particular concern as they can
penetrate deep into the lungs, and their exposure has been associated
with asthma, heart disease, heart failure, stroke, and cancer
\citep{mohapatra2014effect, xing2016impact}. As a result, the health
impacts of ambient air particle pollution may be significant, especially
for the most vulnerable. For instance, the \citet{ihme2019ambient}
estimates PM related air pollution resulted in 4.14 million deaths
worldwide (in 2019), of which, 394,000 occurred in Africa. Yet, despite
its importance, the air quality and health picture in Africa is
worryingly incomplete. For instance, the WHO only collects data from 10
African countries (covering 39 cities)
\citep{worldhealthorganization2016ambient}. Likewise, robust,
epidemiological data from and for the African context is also lacking,
though emerging data suggests strong correlations between PM and poor
health outcomes \citep{coker2018narrative, heft-neal2018robust}. Moving
forward, it is not just the air at Queen's which lacks clarity.

@@

Findings suggest that particulate matter concentrations are routinely
above the WHO guidelines and are especially worrisome at several
locations. Over the course of the two-month study, the hazardous limits
for both PM\textsubscript{10} and PM\textsubscript{2.5} were exceeded at
all locations except for the Lions Sight hospital. The limits were
exceeded fewer than 50 times at five of the locations, but the monitors
at the Lighthouse Clinic and the Guardian Shelter recorded more than 50
instances above the hazardous limits. The extremely hazardous air
quality at the Guardian shelter is mostly from cooking over open fires,
while the PM at the Lighthouse Clinic is a function of being located
directly adjacent to the smouldering waste pile.

@@

Yet, according to respondents and our own observations, waste management
procedures within the hospital are fraught with challenges which impact
the implementation of best practices. These include the lack of a
coherent hospital-wide waste management policy, insufficient waste
management training for staff, and a persistent lack of material and
financial resources. For instance, although hospital administration was
able to articulate a set of standard operating principles (SOPs) for
waste management, below the administration, doctors and nurses were
generally limited in their understanding about its contents.
Furthermore, cleaners, who are responsible for day-to-day cleansing and
waste disposal, yet rate at the bottom of the institutional hierarchy at
Queens, were unable to access or analyse hospital policy, and generally
tended to rely on ``experience'' and ``human judgement'' when handling
potentially hazardous waste. Moreover, although nearly all doctors and
nurses had prior education on waste management best practices, cleaners
began their waste handling duties without any prior waste management
experience, and although there is an orientation programme for new
hires, there is no regular refresher training or capacity building
regarding waste for existing staff, at any level.

@@

The PM monitoring device centres around a Raspberry Pi 3 Model B
single-board computer complete with an operating system and storage
space. The added advantage of storage space is that the data are not
lost if the internet connection breaks down. Because the Pi can only
read digital signals, we needed to include an analogue-to-digital (ADC)
converter. The amount of dust is measured by a Nova PM SDS011 High
Precision Laser sensor. This sensor measures particles at 2.5 and 10 µm
in diameter in mg/m\textsuperscript{3}. A list of the specific hardware
and software components can be found in Appendix B.

@@

\hypertarget{computational-reproducibility-and-data-availability}{%
\subsubsection{Computational reproducibility and data
availability}\label{computational-reproducibility-and-data-availability}}

R Statistical Software version 4.2.1, RStudio IDE version 2023.3.0.386,
and Quarto scientific publishing system version were used for
quantitative data analysis and writing of the manuscript
\citep{R-base, allaire2022quartoa, positteam2023rstudio}. A set of
additional R packages were used for data wrangling, analysis, and
visualisation
\citep{ggplot22016, lubridate2011, R-dplyr, R-forcats, R-ggplot2, R-janitor, R-leaflet, R-lubridate, R-mapview, R-readr, R-tidyr, R-waffle}.

Raw data and analysis-ready processed data is available as an R package
at \citet{R-qechairquality}. The data underlying the tables and figures
of this manuscript are contained in a repository alongside a
reproducible document that contains the analysis code and the written
narrative of the manuscript {[}TODO: manuscript software citation{]}.

@@

The Malawi Bureau of Standards is the national agency responsible for
setting and publishing all standards in the country. At the time of
writing, the official website (www.mbsmw.org) was not available.
However, published work that references Malawian standards
\citep{kutlarjoss2017time, mapoma2014air} indicate that the maximum 24-h
PM\textsubscript{10} value is 25 µg/m\textsuperscript{3} and the maximum
annual PM\textsubscript{2.5} value is 8 µg/m\textsuperscript{3}. There
is no daily maximum value for PM\textsubscript{2.5}. It should be noted
that both of these values are at, or below, the level-4 interim WHO
targets.

@@

\hypertarget{tbl-peaks}{}
\begin{longtable}[]{@{}lrrr@{}}
\caption{\label{tbl-peaks}Peaks for both PM\textsubscript{2.5} and
PM\textsubscript{10} are the number of data points above the WHO interim
target 1 (annual) which is the least stringent; the number of
observations recorded (count) is provided for reference.}\tabularnewline
\toprule()
location & n & pm10 & pm2.5 \\
\midrule()
\endfirsthead
\toprule()
location & n & pm10 & pm2.5 \\
\midrule()
\endhead
Administration & 10900 & 171 & 0 \\
Guardian Shelter & 11032 & 4529 & 0 \\
Lighthouse Clinic & 11472 & 4365 & 0 \\
Lions Sight & 11749 & 91 & 0 \\
Malaria & 12031 & 617 & 0 \\
Mercy James & 12193 & 736 & 0 \\
Ward 6B & 11861 & 1113 & 0 \\
uMoyo & 10637 & 703 & 0 \\
\bottomrule()
\end{longtable}

Over the course of the two-month study, the hazardous limits for both
parameters were exceeded at all locations except for the Lions Sight
hospital. The limits were exceeded fewer than 50 times at five of the
locations, and only the monitors at the Lighthouse Clinic and the
Guardian Shelter recorded more than 50 instances above the hazardous
limits. At the Lighthouse Clinic, the PM\textsubscript{10} limit was
exceeded almost twice as often as the PM\textsubscript{2.5} limit, while
at the Guardian Shelter, the reverse is true. Because the Lighthouse
Clinic is within 50 m of the incinerator, the number and predominance of
PM\textsubscript{10} peaks is characteristic of incomplete combustion
and dust that is typical for the area and the open burning that occurs
there. The residents at the Guardian Shelter however are mostly cooking
within contained clay stoves that have been designed to improve
combustion of the coal and wood that is burned within them. The fewer
number of PM\textsubscript{10} peaks is a testament to this
intervention, though the frequency of PM\textsubscript{2.5} peaks is
still beyond acceptable.

@@

The 24-hour averaged values (logged every 5 minutes) are presented for
each location, for both PM\textsubscript{10} and PM\textsubscript{2.5}
in fig-24-hour-average. As well as dampening local variation and peaks,
the WHO air quality guidelines are also based on 24-hour averages which
are the standard against which the health risks can be judged.

Overall, PM\textsubscript{2.5} values remained below 100
µg/m\textsuperscript{3} at 6 of 8 locations (Administration, Lions
Sight, Malaria Project, uMoyo, Ward 6B, and Mercy James);
PM\textsubscript{10} values were consistently below
100µg/m\textsuperscript{3} at the same locations, but with several
average values extending slightly above, and then infrequently.

@@

\caption{\label{fig-24-hour-average}Average 24-hour PM\textsubscript{10}
and PM\textsubscript{2.5} at 8 monitoring stations over 2 months}

@@

Given that the type of fuel and the type of burning (contained vs.~open)
produce very different particulate ``fingerprints'', the ratio of
PM\textsubscript{10} to PM\textsubscript{2.5} values were examined to
determine if a clear difference between locations, and therefore source,
could be identified. The results are presented in
Figure~\ref{fig-pm-ratios}.

@@

\caption{\label{fig-pm-ratios}PM\textsubscript{10}:PM\textsubscript{2.5}
values for each location. Each line shows the minimum and maximum ratio
for the day based on ratios calculated hourly. Does not include 14th to
16th October at location uMoyo due to extreme outliers.}

@@

The results presented in Figure 5 illustrate both the presence of
PM\textsubscript{10} relative to PM\textsubscript{2.5} as well as the
range of values over which that ratio is observed.

Unlike the peak values presented in Table 2, the values presented in
Figure 5 indicate the relative presence of larger (PM\textsubscript{10})
particles to smaller ones (PM\textsubscript{2.5}) regardless of their
concentration: the ratio of two low concentrations can be the same as
the ratio of two high concentrations and is therefore more indicative of
source than of proximity.

@@

The air quality, as a result of open waste burning at Queen's is poor
and not suitable for a city, let alone a hospital. Over the course of
this two-month study, the hazardous limits for both PM parameters were
exceeded at all locations except for the Lions Sight hospital. The WHO
limits were exceeded fewer than 50 times at five of the locations, and
only the monitors at the Lighthouse Clinic and the Guardian Shelter
recorded more than 50 instances above the most hazardous limit. The
composition of emissions recorded ranged significantly, analysed as the
ratio PM\textsubscript{10}:PM\textsubscript{2.5} likely reflects a range
of burning styles and/or fuel (waste) type, and no discernible trend was
observed.. Among interview respondents, there was a general consensus
that air quality caused by the burning of waste within the hospital was
a problem, although there were significant differences between
respondents over their understandings of potential impacts, the
effectiveness of various coping mechanisms, and their problematisations
linked to the burning of specific waste materials. For most, going home
or leaving the Queen's campus was the only pathway to relief, though for
patients who can not leave, endurance was the only option. This work was
originally planned as a before-and-after study: we envisioned reporting
these baseline results compared to follow-up measurements once the
long-promised incinerator was commissioned. However, nearly 4 years
later, the incinerator is rarely functional (partly for technical
reasons, partly for financial ones) and the smoke hovering over Queen's
is still thick. Waste management has long been known as an environmental
health issue, but in the case of open trash burning at a medical
facility, it becomes more: for some of the most vulnerable, especially
those with TB, asthma or other chronic disease, it could be a matter of
life or death.

@@

\caption{\label{fig-five-minute-intervals}PM\textsubscript{2.5} and
PM\textsubscript{10} values collected every 5 minutes over 10 months at
8 locations.}

@@

\begin{itemize}
\tightlist
\item
  analogue-to-digital converter (ADC)
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  any box that can fit all the components comfortably can be used. The
  box will need to have a whole drilled in it for the pipe.
\end{itemize}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The Raspberry Pi operating system runs off of Rasbian software on a
  NOOBS scandisk.
\item
  Air quality monitoring code is written in Python using the
  \href{https://pypi.org/project/python-aqi/}{aqi library} (A library of
  algorithms to convert between AQI value and pollutant concentration).
  The code is stored on the
  \href{https://github.com/opendatadurban/hospital_stations}{Open Data
  Durban GitHub account} and is free for all to access. To be able to
  collect data, the device needs to be connected to a reliable wifi
  network, the SSID and password need to be coded onto the device during
  step 2 in the building method.
\item
  Once the whole device has been put together and placed on site, to
  access that data we use a combination of programs:
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set up the Raspberry Pi with the Rasbian software using the NOOBS
  scandisk. (insert the scandisk into the Raspberry Pi) A comprehensive
  start up guide can be found here.
\item
  Next we load the code onto the Raspberry Pi. The code can be found on
  the Open Data Durban Hospital Stations Repository Github. To use it
  you can follow the steps in the file called Initiate\_WS.txt. Once all
  the steps have been completed, the Raspberry Pi is now a device that
  can be connected to a particulate matter monitoring sensor.\\
\item
  Connect the hardware components
\end{enumerate}
